\documentclass{ximera}
\title{Functions of Two Independent Variables, Introduction}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Introduction to traces, level curves and surfaces.
\end{abstract}
\maketitle


\begin{example} \label{ExMD3f039}

The function
\[
   z = f(x,y)
\]
gives the temperature (in Celsius degrees) at the point in the plane with coordinates $(x,y)$ (measured in meters).

Some level sets of the function $f$ are graphed below.

\begin{onlineOnly}
    \begin{center}
\desmos{ve90qgcwoh}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ve90qgcwoh}{163: Level Sets 1}

\begin{enumerate}
\item Explain the meaning of the level set $f(x,y)=4$.

\item Drag the slider $k_2$ in Line 4 to see how the level set $f(x,y)=k_2$ changes as the temperature $k_2$ changes. 

\item Find the range of the function $z=f(x,y)$, Use set-builder notation.

\item Find the domain of the function $z=f(x,y)$. Use set-builder notation.

\item Zoom in on the point $P(2,-2)$ to approximate the partial derivative
\[
      \frac{\partial f(x,y)}{\partial x}\Big|_{(x,y)=(2,-2)}.
\]
Include units. Then use the language of small changes to interpret its meaning.

Keep in mind the level sets are spaced $1^\circ$C apart and that the green tick marks are spaced $0.20$ meters apart.

\item Drag the slider $\alpha$ in Line 19 to $\alpha=\pi/2$. Then approximate the partial derivative
\[
          \frac{\partial f(x,y)}{\partial y}\Big|_{(x,y)=(2,-2)} .
\]
Include units. Use the language of small changes to interpret its meaning.

\item Now use the fact that
\[
    z = f(x,y) = 10-x^{2}-4xy-4y^{2}
\]
to compute the exact values of the two partial derivatives above.

\item Try to sketch the graph of the function $z=f(x,y)$.


\end{enumerate}

\end{example}


\begin{example} \label{ExLdkjr3rffds}

Some level sets of the function $z=f(x,y)$ are shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{avxpstt5t2}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/avxpstt5t2}{163: Level Sets 1}

Assume $f(x,y)$ is either strictly increasing or decreasing between contour lines.

\begin{enumerate}
\item At which of the labeled points is
\[
    \frac{\partial f(x,y)}{\partial x} > 0 \text{ and }  \frac{\partial f(x,y)}{\partial y} = 0 ?
\]
Then approximate 
\[
  \frac{\partial f(x,y)}{\partial x}
\]
at these points. Explain your reasoning.

\item At which of the labeled points is
\[
    \frac{\partial f(x,y)}{\partial x} = 0 \text{ and }  \frac{\partial f(x,y)}{\partial y} < 0 ?
\]
Then approximate
\[
  \frac{\partial f(x,y)}{\partial y}
\]
at these points. Explain your reasoning.

\item Turn off the folder in Line 1. Then activate the folder in Line 3 and repeat parts (a) and (b).
\end{enumerate}

 
\end{example}



\begin{example}  \label{Edfhf45}
\begin{enumerate}
\item Move the slider $k$ to see the level curves 
\[
    f(x,y)=k
\]
of the function $z=f(x,y)$. Use the level curves to sketch the surface $z=f(x,y)$. Then check the \emph{Surface} folder in Line 8 to see how you did. Change the setting to \emph{Transparent Surfaces} by clicking the wrench in the upper right corner of the worksheet and selecting \emph{More Options}.


\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/3d/u5syjyxcsw}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmosThreeD{u5syjyxcsw}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/u5syjyxcsw}{163: Hemisphere}

\item Deactivate the folder \emph{Surface} in Line 8. Then use the level curves of the function $z = f(x,y)$ above to approximate the partial derivative
\[
    \frac{\partial z}{\partial x} \Big|_{(x,y)= (-2,1)} .
\]
Keep in mind the spacing of the level curves (Line 6) in computing the approximation.

\item Drag the slider $\alpha$ in Line 7 to approximate the partial derivative
\[
    \frac{\partial z}{\partial y} \Big|_{(x,y)= (-2,1)} .
\]

\item Use the expression
\[
  z = f(x,y) = \sqrt{9-x^2}
\]
for the function $f$ to answer the following questions.

\begin{enumerate}
\item Write the domain and range of the function $z=f(x,y)$ in set-builder notation.

\item Evaluate the partial derivatives 
\[
    \frac{\partial z}{\partial x} \Big|_{(x,y)= (-2,1)} 
\]
and
\[
    \frac{\partial z}{\partial y} \Big|_{(x,y)= (-2,1)} 
\]
and compare these with your estimates.

\item Sketch the curve on the surface $z=f(x,y)$ that lies in the plane $y=1$. Then activate the folders \emph{Surface} (Line 8) and \emph{Trace} (Line 10) to see how you did. But first drag the slider $\alpha$ in Line 7 back to $\alpha = 0$.

\item Parameterize the curve in part (iii). Then parameterize the tangent line to the curve at the point $(-2,1,2)$.

 %Then use geometry and also the methods of Calculus 1 to find the slope of the tangent to the cross-section at the point $(-2, 1, 2)$. 

%Then activate the folders \emph{Surface} (Line 8) and \emph{Trace} (Line 10) to see how you did. But first drag the slider $\alpha$ in Line 7 back to $\alpha = 0$.

\item Sketch the curve on the surface $z=f(x,y)$ that lies in the plane $x=-2$. Then drag the slider $\alpha$ in Line 7 to $\alpha = \pi/2$ to see how you did.

\item Parameterize the curve in part (v). Then parameterize the tangent line to the curve at the point $(-2,1,2)$.

\item Find an equation of the tangent plane to the surface at the point $(-2,1,2)$.

\end{enumerate}
\end{enumerate}
\end{example}



\begin{example}  \label{Ed4fgvvvvf45}
Move the slider $k$ to see the level curves 
\[
    f(x,y)=k
\]
of the function $z=f(x,y)$. Use the level curves to imagine or draw the surface $z=f(x,y)$. Then check the \emph{Surface} folder in Line 8 to see how you did. Change the setting to \emph{Transparent Surfaces} by clicking the wrench in the upper right corner of the worksheet and selecting \emph{More Options}.


\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/3d/u5syjyxcsw}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmosThreeD{yj0kpv4zmi}{900}{600}
\end{center}
\end{onlineOnly}


 
\href{https://www.desmos.com/3d/yj0kpv4zmi}{163: Hyperboloid}
\end{example}



\begin{example}  \label{Ed44mlk45}
Move the slider $k$ to see the level curves 
\[
    f(x,y)=k
\]
of the function $z=f(x,y)$. Use the level curves to imagine or draw the surface $z=f(x,y)$. Then check the \emph{Surface} folder in Line 8 to see how you did. Change the setting to \emph{Transparent Surfaces} by clicking the wrench in the upper right corner of the worksheet and selecting \emph{More Options}.


\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/3d/u5syjyxcsw}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmosThreeD{vgyxfvgsyp}{900}{600}
\end{center}
\end{onlineOnly}


 
\href{https://www.desmos.com/3d/vgyxfvgsyp}{163: Cone}
\end{example}




\end{document}