\documentclass{ximera}
\title{The Gradient Vector}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
The gradient vector and its properties
\end{abstract}
\maketitle

\section{Working in Two Dimensions}

\begin{question}  \label{Qert4thbhg}
Suppose $z=f(x,y)$ is a differentiable function that gives the temperature (in Celsius degrees) at the point with coordinates $(x,y)$, measured in meters.

Suppose that 
\[
    \Big|  \nabla f(3,5) \Big| = 2 .
\]

and that the vector ${\bf v}$ makes the angle $\theta$ with $\nabla f(3,5)$.

(a) Find an expression for the rate at which the temperature changes at $P$ in the direction of ${\bf v}$. Give a geometric explanation for this expression. No computations.

(b) Suppose the temperature at another point $Q$ increases at the rate of $3^\circ$C/m in the direction of the vector $\langle 24,7\rangle$ and decreases at the rate of $3^\circ$C/m in the direction of the vector $\langle -3,-4\rangle$. Use vector arithmetic (not algebra) to determine the direction in which the temperature at $Q$ increase at the fastest rate. Compute this maximum rate.

\end{question}


\begin{question}  \label{Qdr4356}
The figure below shows two level curves of the differentiable function $z=f(x,y)$ and the gradient of that function at point $C$. Sketch the gradient at point $D$. Explain your reasoning.

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/zau4grcm}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{zau4grcm}{900}{600}
\end{center}
\end{onlineOnly}

Geogebra activity available at

\href{https://www.geogebra.org/classic/zau4grcm}{163: Gradient Exercise}
\end{question}


\begin{question} \label{Q34gvbty}
The function 
\[
 \sigma = f(x,y)
\]
gives the charge density (in $\mu C/m^2$) at the point with $(x,y)$ (measured in meters) on the surface of a thin conducting plate.

Two level curves of the function $f$ are shown below.

\pskip

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/xnqmhyvd}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{xnqmhyvd}{900}{600}
\end{center}
\end{onlineOnly}

Geogebra activity available at

\href{https://www.geogebra.org/classic/xnqmhyvd}{163: Gradient Exercise 2}


Suppose point $C$ has coordinates $(2,5)$,
\[
    f(2,5) = 4,
\]
and that
\[
    \Big|  \langle \frac{\partial z}{\partial x}, \frac{\partial z}{\partial y} \rangle \Big|_{(x,y) =(2,5)} = 30
\]

\pskip

The gradient vector at $C$ is shown in the figure above. Consecutive marks along the line $CD$ are at intervals of length $1$cm. Answer the following questions \emph{without} imposing a coordinate system. Work with the vectors themselves, not with their components.

(a) What are the units of the above gradient? Explain its meaning.

(b) Approximate the charge density at $D$.

(c) Approximate the charge desity at $F$.



\end{question}


\section{Avoiding Implicit Differentiation}

\pskip

Set $z$ free

when wishing to find

a vector normal

to the surface

$f(x,y,z)=9$.

Don't differentiate implicitly,

kiss the chain rule goodbye,

just let 

$w=f(x,y,z)$

and forget about the nine.

$z$ is now free,

it's so simple you see,

just take the partials of $w$

with respect to $x$, $y$, and $z$.

Then you'll have found,

without spending much time,

a vector normal to the surface

$f(x,y,z) = 9$.

What if 

$f(x,y,z) = 10$?

Then do the same thing

all over again.


\pskip\pskip

\begin{question} \label{Qdfbtt44}
Find an equation of the tangent plane to the surface
\[
    xyz - yz^2 +3x^2 -y^2 = 2
\]
at the point $(1,1,1)$.
\end{question}

\begin{question}  \label{Q435r4gt}
(a) Prove that the volumes of the tetrahedra bounded by the tangent planes to the surface
\[
    xyz = k
\]
and the coordinate planes are all equal.

(b) State and prove an analogous theorem in two-dimensions.
\end{question}


\begin{question}  \label{Qweregtt5}
A light at the point $(a,b,c)$ illuminates part of the paraboloid
\[
  k z = x^2 + y^2.
\]
Here $k>0$ is a constant.

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/nfkajvkp}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{nfkajvkp}{900}{600}
\end{center}
\end{onlineOnly}

Geogebra activity available at

\href{https://www.geogebra.org/classic/nfkajvkp}{163: Terminator}



(a) Find an algebraic inequality that gives necessary and sufficient conditions for the point $(x,y,z)$ on the paraboloid to be illuminated.

(b) Parameterize the curve on the surface (the terminator) that forms the boundary separating the illuminated and dark parts of the paraboloid. 

(c) Compute the area of the illuminated part of the paraboloid.
\end{question}


\end{document}