\documentclass{ximera}
\title{Motion: Conceptual Questions}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Conceptual questions about motion. 
\end{abstract}
\maketitle



\begin{question}  \label{QPf433445r}
Play the motion below and count the number of turning points in the graph of the speed function. 

\begin{onlineOnly}
    \begin{center}
\desmos{zce2pvchcg}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/zce2pvchcg}{163: Left or Right}

\end{question}


\begin{question} \label{QPflefemmm}

An ant crawls along the path shown below. Its velocity and acceleration vectors are shown at the moment it passes point $P$ below.

\begin{onlineOnly}
    \begin{center}
\desmos{5ohvr2gp08}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/5ohvr2gp08}{163: Twice as Fast}

A beetle crawls along the same path in the \emph{opposite} direction. It crawls twice as fast as the ant (ie. the beetle passes each point of the path with twice the speed the ant had at that same point).

\begin{enumerate}
\item Sketch the beetle's velocity vector at point $P$.

\item Sketch the beetle's accleration vector at point $P$. Explain your reasoning.

\item Play the slider $u$ in Line 1 to see how you did.

 
\end{enumerate}


\end{question}


\begin{question} \label{Q8f93r3}

Let ${\bf p}(t)$ be a motion in space (with output measured in meters, input measured in seconds). 

Let ${\bf v}(t)$ be the velocity of the motion and ${\bf a}(t)$ its acceleration.

Interpret the meaning of each of the following expressions. Include units.

\begin{enumerate}

\item $\int_3^t {\bf v}(u) \, du$

\item $\int_3^t \left| {\bf v}(u) \right| \, du$

\item $\int_3^t {\bf a}(u) \, du$

\item $\int_3^t \frac{{\bf v}(u)\cdot {\bf p}(u)}{| {\bf p}(u) |} \, du$

\item $\int_3^t \frac{{\bf a}(u)\cdot {\bf v}(u)}{| {\bf v}(u) |} \, du$

\end{enumerate}

\end{question}


\begin{question} \label{Q8dfdff93r3}

Let ${\bf p}(t)$ be a motion in the $xy$-plane of three-dimsional space (with output measured in meters, input measured in seconds). 

Let ${\bf v}(t)$ be the velocity of the motion and ${\bf a}(t)$ its acceleration.

Let ${\bf k} = \langle 0, 0, 1 \rangle$ be the unit vector in the direction of the positive $z$-axis.

Interpret the meaning of each of the following expressions. Include units.

\begin{enumerate}

\item $\frac{\left( {\bf p}(t) \times {\bf v}(t) \right)\cdot {\bf k}}{\left|  {\bf p}(t) \right|^2}$

\item $\int_3^t  \frac{\left( {\bf p}(u) \times {\bf v}(u) \right)\cdot {\bf k}}{\left|  {\bf p}(u) \right|^2}   \, du$

\end{enumerate}

\end{question}





\end{document}