\documentclass{ximera}
\title{Cylinders}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Writing equations of cylinders.
\end{abstract}
\maketitle


\section{The Closest Point on a Line}

\begin{question} \label{QLFKerdfeMdfR}
\begin{enumerate}
\item Parameterize the line through the points $A(-2,1,3)$ and $B(-1,3,1)$.

\item Use the chain rule to find an expression for the derivative
\[
      \frac{d}{dx} \left(  (5x - 3)^2 \right).
\]

\item Use calculus to find the coordinates of the point on the line closest to the point $P$ with coordinates $(6,8,5)$. \emph{Hint:} Treat this as an optimiation problem.

\item Use the worksheet below to check your work.
\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{pcav8vrwva}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/pcav8vrwva}{163: Closest Point on a Line 2}
\end{question}



\section{Right Circular Cylinders in $\mathbb{R}^3$}

A right circular cylinder in $\mathbb{R}^3$ is the set of points a fixed distance from a given line. The line is the cylinder's axis of symmetry. In this section we'll find equations of these cylinders.

\begin{question}  \label{QPdfeerld}

This question is about finding an equation of the right circular cylinder of radius $5$ symmetric about the $y$-axis. This cylinder is the set of points in space exactly $5$ units from the $y$-axis.

\begin{enumerate}
\item Use common sense to find the coordinates of the point $Q$ on the $y$-axis closest to the point $P$ with coordinates $(a,b,c)$.

\item Use the result of part (a) to find an expression for the distance from the point $P$ to the $y$-axis.

\item Use the result of part (b) to find an equation for the cylinder of radius $5$ symmetric about the $y$-axis.

The point $P$ with coordinates $(x,y,z)$ lies on this cylinder if and only if the distance from $P$ to the $y$-axis is equal to $5$, or if and only if
\[
        \answer{\sqrt{x^2 + z^2}} = 5 .
\]

\end{enumerate}
\end{question}

\begin{question}  \label{QLFfr3r3fsdf3}
This question is about find an equation of a right circular cylinder of radius $r_1$ symmetric about the line ${\cal L}$ through the origin parallel to the vector ${\bf w}=\langle w_1,w_2,w_3 \rangle$. The cylinder is the set of points in space exactly $r_1$ units from ${\cal L}$.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{bgdw2yzqsl}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/bgdw2yzqsl}{163: Cylinder}


\begin{enumerate}
\item Use calculus to find an expression for the coordinates of the point $Q$ on ${\cal L}$ closest to the point $P$ with coordinates $(a,b,c)$.

\item Enter your expression from part (a) in Line 11 of the worksheet above.

\item Parameterize segment $\overline{PQ}$ in Line 13. Check to see if $Q$ looks correct. Move the sliders $w_1$, $w_2$, $w_3$, $a$, $b$, $c$.

\item Use the result of part (a) to find a equation for the cylinder of radius $r$ symmetric about ${\cal L}$. Enter this equation in Line 14.
\end{enumerate}

\end{question}



\end{document}