\documentclass{ximera}
\title{The Scalar Product}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
An introduction to vector and scalar projections
\end{abstract}
\maketitle

\section{Vector Projections}

It often helps in applications to resolve a vector ${\bf v}$ into two components, one parallel to a given vector ${\bf w}$ and the other perpendicular to ${\bf w}$. For example, resolving the acceleration vector into components parallel and perpendicular to the velocity vector gives information about the external forces acting on an object and how they affect the object's speed and the rate at which it changes direction.


\begin{exploration}
Drag points $A$, $B$, $D$, and $E$ to see how the vector projection ${\bf v}_{\text{par}}$ of ${\bf v}$ in the direction of ${\bf w}$ changes. Watch also the vector component ${\bf v}_{\text{perp}}$ of ${\bf v}$ perpendicular to ${\bf w}$.


\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/calculator/dfwwd9dg}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{dfwwd9dg}{900}{600}
\end{center}
\end{onlineOnly}
\end{exploration}





The \emph{vector projection} of the vector ${\bf v}$ in the direction of the vector ${\bf w}$ is the vector ${\bf v}_{par}$ parallel to ${\bf w}$ that makes the difference
\[
      {\bf v}_{perp} = {\bf v} - {\bf v}_{par}
\]
perpendicular to ${\bf w}$.

In two dimensions slope makes it easy to determine if two vectors are parallel. But we cannot really talk about the slope of a vector in $\mathbb{R}^3$ or higher dimensions. So to resolve a vector into two components as just described, we need a way to determine whether two vectors in the same dimension are perpendicular. It you think about this problem from scratch (ie. ignoring what you may have read in another source), the solution is not immediately obvious.

\begin{question}  \label{Q235r74:Scalar}
Which of the following are equivalent to the condition that the non-zero vectors ${\bf a}, {\bf b} \in \mathbb{R}^n$ are perpendicular? Check all that apply. Draw pictures to explain your reasoning. No need to compute.

\begin{selectAll}  
    \choice{The vectors ${\bf a}+{\bf b}$ and ${\bf a}-{\bf b}$ are perpendicular.}  
    \choice{The vector ${\bf a}+{\bf b}$ bisects the angle between ${\bf a}$ and ${\bf b}$.}  
    \choice[correct]{$|{\bf a}+{\bf b}| = |{\bf a}-{\bf b}|$}  
    \choice[correct]{$|{\bf a}+{\bf b}|^2 = |{\bf a}|^2 + |{\bf b}|^2$}
      \choice{$|{\bf a}-{\bf b}|^2 = |{\bf a}|^2-|{\bf b}|^2$}
    \choice[correct]{$|{\bf a}-{\bf b}|^2 = |{\bf a}|^2+|{\bf b}|^2$} 
  \end{selectAll}  
 \end{question}


\begin{question}   \label{Q32erdf:Scalar}
Use one of the correct conditions above to find an algebraic condition for the non-zero vectors
\[
    {\bf v}_1= \langle x_1, y_1, z_1 \rangle, {\bf v}_2 = \langle  x_2, y_2, z_2\rangle \in \mathbb{R}^3 
\]
to be perpendicular.

\begin{explanation}
It doesn't matter which of the three equivalent conditions we use, but for reasons of symmetry, we'll go with the first. Then ${\bf v}_1$ and ${\bf v}_2$ are perpendicular if and only if
\[
   |  {\bf v}_1 + {\bf v}_2 | = | {\bf v}_1 - {\bf v}_2 | ,
\]
or in terms of components,
\[
        | \langle   x_1+x_2 , y_1+y_2, z_1 + z_2   \rangle  | = | \langle   x_1-x_2 , y_1-y_2, z_1 - z_2   \rangle  | .
\]
Squaring both sides gives the equivalent condition that
\[
          (x_1 + x_2)^2 + (y_1 + y_2)^2 + (z_1 + z_2)^2 = \answer{(x_1 - x_2)^2 + (y_1 - y_2)^2 + (z_1 - z_2)^2} .
\]

After some algebra, we find that the non-zero vectors ${\bf v}_1$ and ${\bf v}_2$ are perpendicular if and only if
\[
      \answer{x_1 x_2 + y_1y_2 + z_1 z_2} = 0.
\]
\end{explanation}
\end{question}

\begin{question}  \label{Qgsdfbe34:Scalar}
Find three vectors in $\mathbb{R}^3$, all pointing in different directions, that are perpendicular to the vector ${\bf v}= \langle -1,3,11\rangle$. 

\begin{explanation}
Using what we just learned, we know a non-zero vector ${\bf w} = \langle x, y, z \rangle \in \mathbb{R}^3$ is perpendicular to ${\bf v} = \langle -1,3,11\rangle$ if and only if
\[
     -x + \answer{3}y + \answer{11}z = \answer{0}.
\]
Now we've turned our geometric problem into an algebraic one. We can just let $y$ and $z$ take arbitrary real values and solve the above equation for $x$. 
\begin{freeResponse}
Carry out the above computations to find three vectors all pointing in different directions, that are perpendicular to ${\bf v}$.
\end{freeResponse}


\end{explanation}
\end{question}



\begin{question} \label{Qdf6:Scalar}
\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/mgihe5dswd}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{mgihe5dswd}{900}{600}
\end{center}
\end{onlineOnly}


\end{question}


\begin{question}   \label{Qrasd5r:Scalar}
(a) Given vectors ${\bf v}, {\bf w} \in \mathbb{R}^n$, use the scalar product to determine the value of the scalar $\lambda\in \mathbb{R}$ that makes the vector
\[
     {\bf v}_{\perp} = {\bf v} - \lambda {\bf w}
\]
perpendicular to ${\bf w}$.

(b) Use part (a) to find an expression for the vector component of ${\bf v}$ in the direction of ${\bf w}$.

\begin{explanation}
The vector ${\bf v}_{\perp}$ is perpendicular to ${\bf w}$ if and only if
\begin{align*}
     0        &=   \textbf{v}_{\perp} \cdot \textbf{w}   \\ 
               &= \left( {\bf v} - \lambda {\bf w} \right) \cdot {\bf w} \\
               &= {\bf v}\cdot {\bf w} - \lambda ({\bf w}\cdot {\bf w}) .
\end{align*}
Solving the above equation for $\lambda$ tells us that
\[
    \lambda = \frac{\answer{{\bf v}}\cdot \answer{{\bf w}}}{\answer{{\bf w}}\cdot \answer{{\bf w}}} .
\]
So the vector component of ${\bf v}$ in the direction of ${\bf w}$ is
\[
    \text{proj}_{{\bf w}}({\bf v}) = {\bf v}_{par} = \left(  \frac{\answer{{\bf v}}\cdot \answer{{\bf w}}}{\answer{{\bf w}}\cdot \answer{{\bf w}}}    \right) \answer{\bf w}.
\]
\end{explanation}
\end{question}


\begin{question} \label{Q98734454331}
Find an equation of the circle of radius $r$ tangent to the curve
\[
    y=10\ln (x/5)
\]
at the point $B(5,0)$ with its center below the curve.

 
\begin{onlineOnly}
    \begin{center}
\geogebra{fgvvdjaf}{900}{600}
\end{center}
\end{onlineOnly}
 
\href{https://www.geogebra.org/classic/fgvvdjaf}{163: Osculating Circle}
 
\end{question}


\begin{question}  \label{Qfgsadge:Scalar}
Let ${\cal L}$ be the line through the origin and the point $A(3,1,4)$. Let $Q$ be the point with coordinates $(11,5,10)$.

(a) Use a vector projection to find the coordinates of the point on ${\cal L}$ closest to $Q$.

(b) Use calculus to find the coordinates of the point on ${\cal L}$ closest to $Q$.

(c) Compare your solutions to (a) and (b).

(d) Find an equation of the circular cylinder through $Q$ that is symmetric about ${\cal L}$. {\it Hint:} Start by descrbing the cylinder as a set of points satisfying some condition involving distance. Then translate this description into an equation.

(e) Find an equation of the circular cone through $Q$ with its vertex at the origin that is symmetric about ${\cal L}$. Start by descrbing the cylinder as a set of points satisfying some condition involving distance. Then translate this description into an equation.
\end{question}


\begin{question}  \label{Qfgs643r6e:Scalar}
Let ${\cal L}$ be the line through the points $A(3,1,4)$ and $B(-3,-5,-8)$. Let $Q$ be the point with coordinates $(7,-3,-2)$.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{lrvizzj9jf}{900}{600}
\end{center}
\end{onlineOnly}
 
\href{https://www.desmos.com/3d/lrvizzj9jf}{163: Vector Projection 1}

(a) Use a vector projection to find the coordinates of the point on ${\cal L}$ closest to $Q$.

(b) Find an equation of the circular cylinder through $Q$ that is symmetric about ${\cal L}$. 

(c) Find an equation of the circular cone through $Q$ with its vertex at $A$ that is symmetric about ${\cal L}$. Start by descrbing the cone as a set of points satisfying some condition involving distance. Then translate this description into an equation.




\end{question}



\section{Light Reflecting off a Mirror}
\begin{exploration}  \label{E32fvsav:Scalar}
Light hits the mirror below moving parallel to the vector ${\bf v}_{i}$. It refects off the mirror heading in the direction parallel to the vector ${\bf v}_o$. The vector ${\bf w}$ is parallel to the mirror.

Express the vector ${\bf v}_o$ in terms of the vectors ${\bf v}_{i}$ and ${\bf w}$.

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/xvxbt43k}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{xvxbt43k}{900}{600}
\end{center}
\end{onlineOnly}


\end{exploration}



\begin{exploration}  \label{E32fvsdf9v:Scalar}

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/1xrumhkojh}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{1xrumhkojh}{900}{600}
\end{center}
\end{onlineOnly}


\end{exploration}



\section{Billiards Anyone?}

\begin{exploration}\label{exp:pc1c}
Play the slider $s$. Control the direction of the cue ball with the slider $\phi$. 
%Enter the motions of the balls in the desmos activity in Lines ??-?? below.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/8p4pp5ri6k}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{8p4pp5ri6k}{900}{600}
\end{center}
\end{onlineOnly}
\end{exploration}


\section{Scalar Projections: Changing Coordinates}
A simple way to think about scalar projections is via a change in coordinates. Given the coordinates of a point $P$ with respect to the $xy$-coordinate system, we wish to determine its coordinates with respect to some other coordinate system with the same origin but with its axes pointing in some other directions.

\begin{exploration}
In the demonstration below point $P$ has coordinates $(-8,5)$ relative to the $xy$-coordinate system. The dotted blue lines show another coordinate system that we'll call the $uv$-coordinate system. The positive $u$-axis points in the direction of the vector ${\bf u}=\langle 4,3\rangle$ and the positive $v$-axis points in the direction of the vector ${\bf v}=\langle -3,4\rangle$. Our problem is to find the coordinates of $P$ in the $uv$-coordinate system.

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/fppzrurp}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{fppzrurp}{900}{600}
\end{center}
\end{onlineOnly}
\end{exploration}


\begin{question}  \label{Qgsr3452:Scalar}
(a) Look at the graph above to approximate the coordinates of $P$ in the $uv$-system. Assume the axes have the same scale as in the $xy$-coordinate system.

(b) To find the coordinates of $P$ in the $uv$-system, first compute the vector projections of the vector ${\bf w}=\overrightarrow{OP}$ from the origin to $P$ in the directions of the vectors ${\bf u}$ and ${\bf v}$. Write these projections in the form
\[
   \text{proj}_{\bf u}{\bf w} = \lambda_u {\bf u}
\]
and 
\[
   \text{proj}_{\bf v}{\bf w} = \lambda_v {\bf v} .
\]

(c) Explain why the scalars $\lambda_u$ and $\lambda_v$ are {\bf not} the $uv$-coordinates of $P$. But then explain how to modify these scalars to get the correct coordinates.

(d) Move point $P$ in the demonstation above to a new point, say $Q$. Then estimate and compute the exact $uv$-coordinates of $Q$.

(e) Compute the scalar products $\overrightarrow{OP} \cdot \overrightarrow{OQ}$ in both the $xy$ and $uv$-coordinate systems. How do these compare? Explain.
\end{question}


\begin{question}  \label{Qgvbrh4:Scalar}
(a) Graph the curve
\[
  \sqrt{x} + \sqrt{y} = 1 .
\]

(b) Use the cosine and sine functions to parameterize the curve. Include a domain in the parameterization.

(c) Use the parametrization from part (b) to parameterize the same curve in the $uv$-coordinate system with positve axes pointing in the directions ${\bf u} = <1,-1>$ and ${\bf v}= <1,1>$.

(d) Eliminate the parameter to find a Cartesian equation $v=f(u)$ of the curve in the $uv$-coordinate system. Include the appropriate domain.

\end{question}



\section{Scalar Projections: Distance from a Point to a Line}
In this next problem, we will use a scalar projection to find the distance from a point to a line in $\mathbb{R}^2$.


\begin{question}   \label{Qdsfg4nvc:Scalar}
Find the distance from the point $P$ with coordinates $(8,11)$ to the line
\[
   {\cal L}: 3x + 4y = 36
\] 
as follows:

(a) First use the graph below to estimate the distance.

(b) Find the slope of the line ${\cal L}$ and use it to find a vector normal to ${\cal L}$. Do not rely on the graph to find this vector.

(c) Compute the distance from $P$ to ${\cal L}$ with the appropriate scalar projection.

(d)  Move $P$ to the other side of the line and compute the distance from this new point to the line. 

(e) Compare the computations in parts (c) and (d). How do they differ?

(f) Going back to the original coordinates $(8,11)$ of $P$, use vector arithmetic (not algebra) to find the coordinates of the point on ${\cal L}$ closest to $P$.
\begin{exploration}

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/ebnf2m9w}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{ebnf2m9w}{900}{600}
\end{center}
\end{onlineOnly}
\end{exploration}
\end{question}


\begin{question}  \label{Qfhghhy:Scalar}
Describe algebraically and geometric the set of points in the plane with the property that the sum of their distances to the lines $y=0$ and $y=2x$ is constant and equal to $10$.
\end{question}



\end{document}