\documentclass{ximera}
\title{Directional Derivatives CW}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Introduction to directional derivatives.
\end{abstract}
\maketitle

\section{Introduction}

We've learned how to compute rate of change of a temperature function $T=f(x,y)$ in the directions of the vectors ${\bf i} = \langle 1, 0\rangle$, and ${\bf j} = \langle 0, 1 \rangle$. This section is about computing the rate of change of the temperature function (with respect to distance) in the direction given by a vector ${\bf v} = \langle v_x ,v_y \rangle$. 

The idea is to approximate the temperature function with a linear function in the neighborhood of a point. So we'll start by assuming the temperature function is linear.

\emph{Note:} To compute directional derivatives, we need not use a temperature function, but it is necessary that the inputs of the function have the same units..


\section{Linear Functions}

\begin{question}  \label{Ee53efehdhy}

Some level curves of a temperature function $T = f(x,y)$ are shown below. The output is measured in Celsius degrees and the input in meters.

\pdfOnly{
Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/htee9xke}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{htee9xke}{900}{600}
\end{center}
\end{onlineOnly}

Access Geogebra interactives through the online version of this text at
 
\href{https://www.geogebra.org/classic/htee9xke}{163: Directional Derivative}.


\begin{enumerate}
\item Find an expression for $f$.
\[
     T = f(x,y) = \answer{2x + 4y}
\]

\item Use the sliders $\theta$ and $s$ to approximate the (instantaneous) rate at which the temperature changes from the origin in the directions of the following vectors. Include units.

Note:

\begin{itemize}

\item The slider $\Delta s$ controls the distance (in meters) between consecutive tick marks on the number line through the origin. 

\item The slider $k$ controls the temperature of the red level curve. That is, the red level curve has equation $f(x,y)=k$.

\item The slider $\Delta k$ controls the spacing of the level curves. That is, $\Delta k$ is the temperature difference between consecutive level curves.  

\end{itemize}

\begin{enumerate}

\item ${\bf v} = \langle 1 ,2 \rangle$, where $\theta \sim 1.1$

\item ${\bf w} = \langle 2, -3\rangle$, where $\theta \sim 5.3$.

\end{enumerate}

\item Use a  difference quotient to compute the exact values of the above rates of change. Then compare these with your estimates. Click the arrow to the lower right for a partial solution.

\begin{expandable}
Because the temperature function is linear, we can avoid limits and compute the rate of change at the origin in the direction of the vector ${\bf v} = \langle 1, 2\rangle$ as a difference quotient.

In moving from the origin to the point $(1,2)$ the average rate at which the temperature changes with respect to the distance travelled is
\[
   \frac{\Delta T}{\Delta s} = \frac{f(1,2) - f(0,0)}{|\langle 1, 2 \rangle|}\frac{^\circ C}{m} = \frac{10}{\sqrt{5}}\frac{^\circ C}{m} .
\]

Then make a similar computation for rate of change in the direction of the vector ${\bf w} = \langle -3,1\rangle$.
\end{expandable}


\item Use a difference quotient to compute the rate  at which the temperature changes from the origin in the direction of the vector ${\bf v}=\langle v_x , v_y \rangle$. 

\item Check that your expression in the previous part has the correct units.

\item Interpret your expression for the rate of change as the scalar projection of one vector onto another. Then write this scalar projection as the product of a magnitude of the gradient vector and a trigonometric function of the angle between the gradient vector and the vector ${\bf v}$.

\end{enumerate}
\end{question}




\section{General Differentiable Functions}


\begin{question} \label{Qmer34}

Suppose that from your current position, the temperature increases at the rate of $0.02^\circ$C/meter in the direction of the vector $\langle 1, 0 \rangle$ and that it increases at the rate of $0.04^\circ$C/meter in the direction of the vector $\langle 0,1\rangle$. What do you think is true about the rate at which the temperature changes in the direction of the vector ${\bf v} = \langle 1,1\rangle$?

\begin{enumerate}

\item The temperature increases at the rate of exactly $0.03^\circ$C/meter.

\item The temperature increases at a rate somewhere between $0.02^\circ$C/meter and $0.04^\circ$C/meter, but not necessarily at the exact rate of $0.03^\circ$C/meter.

\item The temperature increases at a rate less than $0.02^\circ$C/meter.

\item The temperature increases at a rate greater than $0.04^\circ$C/meter.

\item The temperature decreases.

\item We cannot say, even if we assume (as do the above choices) that the temperature is a differentiable function of position.

Try to justify your response with an intuitive argument without doing any computations.
\begin{freeResponse}
\end{freeResponse}

\item Compute the exact rate of change to see if you were correct.
\end{enumerate}

\end{question}


\begin{question} \label{Qdlf3r3r3}

Some level curves of a temperature function $T = f(x,y)$ are shown below. The output is measured in Celsius degrees and the input in meters.


\begin{onlineOnly}
    \begin{center}
\desmos{s1u2zdsq99}{900}{600}
\end{center}
\end{onlineOnly}

Access desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/s1u2zdsq99}{163: Linear Approximation Level Sets 2}.

\begin{enumerate}
\item Drag the slider $k$ to zoom in on the point $(1,1)$. Then estimate the rate of change of temperature with respect to distance at the point $(1,1)$ in the direction of the vector ${\bf v}=\langle -2, -1 \rangle$. Include units.

\item Use the fact that 
\[
  T  = f\left(x,y\right)=x^{2}-0.5xy^{2}+y^{3}
\]
to compute the exact rate of change in part (a).

\item Find the exact rate of change in the temperature with respect to distance at the point $(2,2)$ in the direction toward the point $(-1,1)$.

\item In what direction from the point $(1,1)$ does the temperature increase at the fastest rate? Find this rate.

\item In what direction from the point $(1,1)$ does the temperature decrease at the fastest rate? Find this rate.

\item In what directions from the point $(1,1)$ does the temperature increase at one-third of its maximum rate? Use vector arithmetic, not algebra to find the direction(s)..

\item Find all directions from the point $(1,1)$ in which the  temperature remains constant.
\end{enumerate}

\end{question}

\section{A Lake Floor}

\begin{question} \label{Q761zd}
Relative to the horizontal, the bottom of a lake directly below your rowboat slopes downward at an angle of $\alpha=\arctan(0.3)$ in the direction due east. The bottom slopes  downward at an angle of $\beta=\arctan(0.2)$ in the direction due east.

Assume the depth of the lake below your boat is a differentiable function of your position.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{16qfpy1rei}{900}{600}
\end{center}
\end{onlineOnly}

Access desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/3d/16qfpy1rei}{163: Lake Floor}

\begin{enumerate}
\item What angle with the horizontal does the bottom make with the horizontal under your boat in the direction of the bearing $\theta_1 = 0.5$ radians? The bearing is measured counterclockwise from the east.

\item In what directions does the lake bottom beneath your boat slope upward?

\item In what direction does the lake bottom beneath your boat slope upward at the steepest angle? Find this angle.

\item If you happened to be passing your current position with a speed of $4$ meters/sec heading directly due northeast, at what rate would the depth of the water beneath your boat have been changing at this instant? 

\item A fish swimming on the bottom of the lake stays directly below your boat at all times. Find the fish's speed at the moment in part (d).
\end{enumerate}

\end{question}

\end{document}