\documentclass{ximera}
\title{Vectors}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Vectors in $\mathbb{R}^2$ and  $\mathbb{R}^3$.
\end{abstract}
\maketitle

\section{Position Vectors}

Position is relative. You describe where something is not in absolute terms but only in relation to something else.

We might say, for example, that Bellevue is 4 miles east and 3 miles north of Akron. Then in a rectangular coordinate system with the $x$-axis pointing due east, the $y$-axis pointing due north, and the coordinates measured in miles, we can represent the position of Bellevue (at point $B$) relative to Akron (at point $A$) by the vector
\[
   \overrightarrow{AB} = \langle 4, 3 \rangle .
\]
This vector tells how to get from Akron to Bellevue. 


Suppose that in our rectangular coordinate sytstem Charleston (point $C$) has coordinates $(-5,-2)$ and Dayton (point $D$) has coordinates $(-1,1)$. Then to get from Charleston to Dayton we would travel east
\[
     ( -1 - (-5) ) \text{ miles} = 4 \text{ miles}
\]
and north
\[
     ( 1 - (-2)) \text{ miles} = 3\text{ miles} .
\]
So we can represent the position of Dayton relative to Charleston by the vector
\[
  \overrightarrow{CD} =       \langle -1 - (-5) , 1- (-2)  \rangle   =  \langle 4, 3 \rangle .
\] 

So the directions that tell us how to get from Akron to Bellevue also tell us how to get from Charleston to Dayton. The
vectors $\overrightarrow{AB}$ and $\overrightarrow{CD}$ are equal and we write 
\[
      \overrightarrow{AB} = \overrightarrow{CD} .
\]

To go the other way and get from Bellevue to Akron or from Dayton to Charleston, we would go 4 miles west and 3 miles south. So the  position of Akron relative to Bellevue, or of Charleston relative to Dayton, is given by the vector
\[ 
   \overrightarrow{BA} = - \overrightarrow{AB} = - \langle 4,3 \rangle =  \langle -4,-3 \rangle.
\]

The distance between Akron and Bellvue is just the length of the vector $\overrightarrow{AB}$, written as
\[
    | \overrightarrow{AB}  | =  | \langle 3, 4 \rangle  | \text{ miles} = \sqrt{3^2 + 4^2} \text{miles} = 5 \text{ miles} .
\] 


\section{Adding and Subtracting Vectors}

\begin{question}    \label{Q234234:Vectors}
Continuing with our example above, suppose that Frankfort (point $F$) is $2$ miles east and $5$ miles south of Bellevue. Then the vector
\[
    \overrightarrow{BF}  = \langle \answer{2}, \answer{-5}\rangle
\]
gives the position of Frankfort relative to Bellvue. 
\end{question}

\begin{question} \label{Q233:Vectors}
(a) Give directions to get from Akron to Frankfort.

(b) Fill in the vector and its components that gives the position of Franfort relative to Akron.
\[
       \overrightarrow{\answer{A} \answer{F}}    =  \langle  \answer{6} , \answer{-1}  \rangle 
\]

(c) Which of the following are correct? Choose all that apply.
\begin{selectAll}  
\choice[correct]{$\overrightarrow{AF}   = \overrightarrow{AB} + \overrightarrow{BF}$}  
\choice{$\overrightarrow{AF}   = \overrightarrow{AB} - \overrightarrow{BF} $}  
\choice[correct]{$\overrightarrow{AF}   = \overrightarrow{BF} + \overrightarrow{AB} $}  
\end{selectAll} 


\end{question}


\begin{question}  \label{Q543:Vectors}
Let ${\bf v}$ be the vector that gives the position of point $K$ relative to point $G$ and ${\bf w}$ the vector that gives the position of $M$ relative to $K$. 

(a) Express in terms of ${\bf v}$ and ${\bf w}$ each of the following vectors: 

(i) The position of $M$ relative to $G$:  $\overrightarrow{\bf \answer{v}} +  \overrightarrow{\bf \answer{w}} $

(ii) The position of $G$ relative to $M$:  $\overrightarrow{\bf \answer{-v}} +  \overrightarrow{\bf \answer{-w}} $

\pskip

(b) Draw vectors ${\bf v}$ and ${\bf w}$ tail-to-tail. Any choices will do, just make sure the vectors are not parallel. Then draw the vector ${\bf v} + {\bf w}$

(c) Explain the meaning of each of the following in the context of this scenario. Include units. 

(i) $|  {\bf v} |$

(ii) $|  {\bf w} |$

(iii) $|  {\bf v+w} |$

(iv) $|  {\bf v} | + |  {\bf w} | - |  {\bf v+w} |$

\pskip

(d) When would the expression in part (b)(iv) above be positive? Negative? Zero? Explain your reasoning. 

\end{question}


\begin{question}  \label{Qsdfsdt4r3:Vectors}
Suppose
\[
 \overrightarrow{AB} = \langle  -3,5 \rangle
\]
and
\[
   \overrightarrow{AC} = \langle  4,2 \rangle ,
\]
where the components are measured in miles.

(a) Give directions to get from $B$ to $C$.

(b) What vector gives the position of $C$ relative to $B$? Choose all that apply.
\begin{multipleChoice} 
 \choice[correct]{$\langle  7, -3  \rangle$}  
 \choice{$\langle  -7, 3  \rangle$}
\choice{$\overrightarrow{CB}$}  
\choice[correct]{$\overrightarrow{BC}$} 
\choice{$\overrightarrow{AB} -\overrightarrow{AC} $} 
\choice[correct]{$\overrightarrow{AC} -\overrightarrow{AB} $}
\end{multipleChoice} 

\end{question}





\begin{question}  \label{Q234r3:Vectors}
Let ${\bf v}$ be the vector that gives the position of point $Q$ relative to point $R$ and ${\bf w}$ the vector that gives the position of $S$ relative to $R$. 

(a) Which of the following express the position of $S$ relative to $Q$? Choose all that apply.

\begin{selectAll}  
\choice{${\bf v} + {\bf w}$}  
\choice{${\bf v} - {\bf w}$}  
\choice[correct]{${\bf w} - {\bf v}$}  
\end{selectAll}

(b) Draw a picture illustrating part (a). Include the vectors ${\bf v}$ and ${\bf w}$ tail-to-tail, and the vector that gives the position of $S$ relative to $R$. That's it. Do not draw a fourth vector like $-{\bf v}$ or $-{\bf w}$.


\end{question}


\begin{question}  \label{Qtete4r3:Vectors}
Let $A$, $B$, and $C$ be non-colinear points. Let ${\bf b}$ be the vector giving the position of $B$ relative to $A$ and let ${\bf c}$ be the vector giving the position of $C$ relative to $A$. 

Suppose between times $t=-24$ and $t=24$ seconds you walk with a constant velocity (ie. with a constant speed in a fixed direction), passing point $B$ at time $t=0$ seconds and point $C$ at time $t=10$ seconds.

Express each of the following in terms of ${\bf b}$ and ${\bf c}$. Draw a picture for each and explain your reasoning.

(i) your position relative to $B$ at time $t=5$ seconds.

(ii) your position relative to $A$ at time $t=5$ seconds.

(iii) your position relative to $C$  at time $t=8$ seconds.

(iv) your position relative to $A$  at time $t=8$ seconds.

(v) your position relative to $A$  at time $t=24$ seconds.

(vi) your position relative to $A$  at time $t=-24$ seconds.

(vii)  your position relative to $B$  at time $t$ seconds.

(vii)  your position relative to $C$  at time $t$ seconds.

(viii)  your position relative to $A$  at time $t$ seconds.

\end{question}





IGNORE THIS

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.www.geogebra.org/classic/e6rsvnsz}.
}
 
\begin{onlineOnly}
    \begin{center}
\geogebra{e6rsvnsz}{900}{600}
\end{center}
\end{onlineOnly}



\section{Computing Coordinates With Vectors}

\begin{question} \label{Q45340:Vectors}
(a) Explain why the spheres with equations
\[
   (x-4)^2 + (y+5)^2 + (z-1)^2 = 49
\]
and
\[
    (x-6)^2 + (y+7)^2 + (z-2)^2 = 100
\]
are tangent to each other.

(b) Use vectors, not algebra, to determine the coordinates of the point of tangency.

(c) Use algebra to determine the coordinates of the point of tangency.

\end{question}


\begin{question} \label{Q45323240:Vectors}
Points $A$ and $B$ have respective coordinates $(5,1,-8)$ and $(7,-2,-14)$. 

(a) Use vectors, not algebra, to determine the coordinates of the point $P$ on segment $\overline{AB}$ that is exactly 5 units from $A$.

(b) Use vectors, not algebra, to determine the coordinates of another point $P$ on the line through $A$ and $B$ that is exactly 5 units from $A$.

(c) Solve questions (a) and (b) simultaneously using algebra, not vectors.

\end{question}


\begin{question} \label{Q4df8240:Vectors}
(a) Use vectors to determine the coordinates of the centers of all circles with radius 6 that are tangent to the ellipse
\[
    x^2 - xy + y^2 = 19
\]
at the point $(2,5)$.

(b) Use the method of part (a) to determine the coordinates of the centers of all circles with radius $r$ that are tangent to the ellipse.

(c) Follow the directions in the demonstration below to check your work.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/aqpedvdawb}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{aqpedvdawb}{900}{600}
\end{center}
\end{onlineOnly}

\end{question}


\begin{question}   \label{Q6547:Vectors}
(a) Use vectors, not trigonometry, to help find equations of the two lines that bisect the angles between the lines $y=x$ and $y=3x$.

(b) Use part (a) to find equations of the two lines that bisect the angles between the lines $y=x+3$ and $y=3x-5$.


(c) Use vectors to help find an algebraic description of the set of points twice as far from the line $y=x$ as from the line $y=3x$. Keep in mind that the distance from a point $P$ to a line is the minimum distance between $P$ and the points of the line. 

\end{question}

\begin{question} \label{Q5704r:Vectors}
The rear wheel of a bicycle moves along the curve
\[
    y = 4 -\int_2^x \frac{2}{\sqrt{5+u^2}}\, du \, , \, -10 \leq x \leq 10 
\]
in the direction of decreasing $x$-coordinates. The coordinates are measure in feet. The front and rear axles are three feet apart.

(a) Find the coordinates of the point where the front wheel touches the $xy$-plane when the rear wheel touches the plane at the point $(2,4)$. 

(b) Enter these coordinates in the desmos demonstration below. 

(c) Find an equation of the segment running between the two contact points and enter it in desmos.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/a20scadjvh}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{a20scadjvh}{900}{600}
\end{center}
\end{onlineOnly}

\end{question}


\begin{question} \label{Q5sdf04r:Vectors}
The purpose of this problem is to find equations of several spheres with a given radius that pass through two given points. Later, we will find equations of all these spheres.

So we'll start with two points $A$ and $B$ with respective coordinates $(2,4,6)$ and $(6,-4,10)$. Our problem is to find three spheres of radius $7$ through $A$ and $B$.

(a) We'll first ignore the condition that the spheres have radius $7$ and consider the family of all spheres through $A$ and $B$.

What can you say about the set of centers of all such spheres? Choose all that apply.
\begin{selectAll}
\choice{The set contains exactly one point, the midpoint of segment $\overline{AB}$}.
\choice{The set is a sphere with diameter $\overline{AB}$}.
\choice[correct]{The set consists of all points equidistant (ie. the same distance) from $A$ and $B$.}
\end{selectAll}


 (b) Using the correct description of the relaxed center-set above and without doing any computations, describe the set geometrically. Be precise. Explain your reasoning.

(c) Translate the correct description of the relaxed center-set in the multiple choice question above directly into an equation of the center-set. Then do some algebra and simplify the equation. Enter it in the line below.
\[
     \answer{x-2y+z=6}
\]

(d) Now let's impose the additional condition that the spheres have radius $7$. Describe geometrically the center-set of all spheres through $A$ and $B$ with radius $7$. Be precise.

(e) Find the coordinates of the center $M$ of the smallest sphere through $A$ and $B$. Use these coordinates and vector arithmetic to find the coordinates of three points that are centers of spheres with radius $7$ through $A$ and $B$. Do this by first finding the coordinates of random points in the relaxed center set (parts (a)-(c)). 

(f) Write equations of three sphere with radius $7$ through $A$ and $B$. Check your work in desmos or geogebra.

\end{question}


\begin{question}  \label{Q45367:Vectors}
Given two spheres that intersect, we'll use vector arithmetic to find the center of their circle of interection.

\begin{itemize}

\item{Let the spheres have centers $(x_1, y_1, z_1)$, $(x_2, y_2, z_2)$ with respective radii $r_1$ and $r_2$.} 

\item{Let ${\bf v}_1$ and ${\bf v}_2$ be the vectors from the origin to the respective centers.}

\end{itemize}

Express the vector ${\bf c}$ from the origin to the center of the circle of intersection in terms of the vectors ${\bf v}_1$, ${\bf v}_2$ and the scalars $r_1$, $r_2$. 

Use desmos or geogebra to check your work. What happens if the spheres do not intersect?

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/q4lnt9y1h4}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{q4lnt9y1h4}{900}{600}
\end{center}
\end{onlineOnly}


\end{question}



\section{Writing Equations with Vectors}

\begin{question}  \label{Eq:32404:Vectors}
In the following questions, we use capital letters like $A$ to denote points and the corresponding boldfaced lower case letter, like ${\bf a}$, to denote the vector from the origin to the point. Also, we use the vector ${\bf r}$ to denote a vector from the origin to the point with coordinates $(x,y)$ in two-dimensions or to the point $(x,y,z)$ in three dimensions.

\pskip

(a) Find a vector equation of the sphere with radius 5 centered at the point $A$.

(b) Find a vector equation of the set of points twice as far from the point $A$ as from $B$.

(c) Find a vector equation of all points the sum of whose distances from $A$ and $B$ is 10 units.


\end{question}






\end{document}