\documentclass{ximera}
\title{Linear Approximation, Introduction}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Introduction to linear approximation.
\end{abstract}
\maketitle


\section{Linear Functions}

\begin{question} \label{Q9f93r33}
A thin rod lies on the $x$-axis between $x=-3$ and $x=3$, coordinates measured in meters.

The function 
\[
      T = f(x) \, , \, -3\leq x \leq 3 ,  
\]
expresses the temperature (in Celsius degrees) of a point on the rod in terms of its $x$-coordinate. Some level sets of the function $T=f(x)$ are shown below. The sets are spaced at intervals of $1^\circ$ C.

\begin{enumerate}
\item Make a reasonable assumption about the function $f$ based on the level sets. Then use your assumption to find an expression for the temperature function $T=f(x)$. Drag point $P$ below to help.
\end{enumerate}


\begin{onlineOnly}
    \begin{center}
\desmos{tokaicoaz7}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/tokaicoaz7}{163: Level Sets Linear Function 2}

\end{question}

\begin{question} \label{Qkdf3r3f00}
The function $T = f(x,y)$ expresses the temperature (in Celsius degrees) of a point in the $xy$-plane in terms of its coordinates (measured in meters).

Some level sets of the function $T=f(x,y)$ are shown below. The level sets are spaced at intervals of $1^\circ$ C.

\begin{enumerate}
\item Approximate the partials 
\[
   \frac{\partial T}{\partial x}\Big|_{(x,y) = (1,2)}
\]
and
\[
   \frac{\partial T}{\partial y}\Big|_{(x,y) = (1,2)}.
\]
Include units.

\item Make a reasonable assumption about the function $f$ based on the level sets. Then use your assumption to find an expression for the temperature function $T=f(x,y)$. Drag point $P$ below to help.
\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmos{9ofxjbrege}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/9ofxjbrege}{163: Level Sets Linear Function}

\end{question}

\begin{question} \label{ExMD3fetet3933}
Hold a rectangular piece of cardboard so that one of its corners rests on a table as shown below.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{zehmvpiqvo}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/zehmvpiqvo}{163: Plane on a Table}

Use vectors to express the height of the opposite corner above the table in terms of the heights of the other two corners.

\end{question}


\begin{question} \label{Q9e8f3rg}
The function
\[
   T = f(x,y)
\]
exrpresses the temperature (in Celsius degrees) of a point in the $xy$-plane in terms of its coordinates (in meters).

Suppose
\begin{enumerate}
\item $f(5,4) = 12$ ,

\item 
\[
      \frac{\partial T}{\partial x}\Big|_{(x,y) = (5,4)} = -2, 
\]
and
\item 
\[
      \frac{\partial T}{\partial y}\Big|_{(x,y) = (5,4)} = 3, 
\]

\end{enumerate}

Now do the following.

\begin{enumerate}

\item State the units of the above partials.

\item Approximate the temperature at the point $(5.2, 4)$. Explain your reasoning.

\item Approximate the temperature at the point $(5, 4.3)$. 

\item Approximate the temperature at the point $(5.2, 4.3)$. Explain your reasoning and how this question is related to Question 3  above and the picture below.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{et3nquim6f}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/et3nquim6f}{163: Linear Approximation 1}


\item Find a linear function $L(x,y)$ that approximates the temperature function near the point $(5,4)$.

\item What additional information might you need to know about the function $f$ to put a bound on the error in your approximation?

\end{enumerate}
\end{question}


\begin{question} \label{Q90f3mmrzx}
The function $T = f(x,y)$ expresses the temperature (in Celsius degrees) of a point in the $xy$-plane in terms of its coordinates (measured in meters).

Some level sets of the function $T=f(x,y)$ are shown below. 

\begin{onlineOnly}
    \begin{center}
\desmos{lyquf7jmbb}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/lyquf7jmbb}{163: Linear Approximation Level Sets}

\begin{enumerate}
\item Drag the slider $k$ in Line 2 to zoom in on the point $(1,1)$. Zoom in close enough so that the level sets look like those of a linear function.

\item Keeping your eye on Line 4 (the spacing of the level sets) and draggin point $P$ to help, approximate partials
\[
   \frac{\partial T}{\partial x}\Big|_{(x,y) = (1,1)}
\]
and
\[
   \frac{\partial T}{\partial y}\Big|_{(x,y) = (1,1)} .
\]
Include units.

\item Now use the fact that 
\[
    T =f(x,y) = x^{2}-.5xy^{2}+y^{3}
\]
to compute the exact values of the partials above. Compare these with your estimates.

\item Find a linear function $L(x,y)$ that approximates the temperature function near the point $(1,1)$. 
\end{enumerate}

\end{question}


\begin{question} \label{QKnDe2111}
This is a continuation of Question 5, where we work in three-dimensions instead of two. We use the same temperature function
\[
    z =f(x,y) = x^{2}-.5xy^{2}+y^{3} .
\]
Its graph is shown below.

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{4urebwqtmk}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/4urebwqtmk}{163: Tangent Plane 2}

\begin{enumerate}
\item Activate Line 1. Then parameterize the curve cut from the surface $z=f(x,y)$ by the plane $y=1$.

\item Activate Line 4. Then parameterize the tangent line to this curve at the point $(1,1,f(1,1))$.

\item Turn off Lines 1 and 4. Then activate Lines 8 and 11. 

\item Parameterize the curve cut from the surface $z=f(x,y)$ by the plane $x=1$ and the tangent line to this curve at the point $(1,1,f(1,1))$.

\item Turn off Lines 8 and 11. Then activate Line 15.

\item Find a vector normal to the tangent plane at $(1,1,f(1,1))$. Then find an equation of this tangent plane. Compare this equation with your answer to Question 5d. Not sure.
\end{enumerate}

\end{question}



\section{The Main Ideas}

Intuititvely, a function $y=f(x)$ of a single variable is \emph{differentiable} at $x=a$ if its graph looks more and more like a non-vertical straight line as we zoom in toward the point $(a,f(a))$. And the slope of that line is the derivative 
\[
   \frac{dy}{dx}\Big|_{x=a} = f^\prime(a).
\]

Somewhat more precisely, a function $y=f(x)$ is differentiable at $x=a$ if the function can be well-approximated by a linear function near $x=a$. Well-approximated in the sense that the error in the approximation approaches zero faster than the change in the $x$-coordinate.

What this means is that we can find a number $m$ such that the linear approximation
\[
   L(x) = f(a) + m(x-a)
\]
is good near $x=a$ in the sense that
\[
   \lim_{x\to a} \frac{f(x) - L(x)}{x-a} = 0.
\]

The value of $m$ is the derivative $f^\prime(a)$ and this definition of differentiability is equivalent to requiring
\[
 \lim_{x\to a} \frac{f(x) - f(a)}{x-a} 
\]
to exist (as a real number). If so, then this limit is by definition the derivative $m=f^\prime(a)$. 

Intuitively, saying a function $z=f(x,y)$ is \emph{differentiable} at $(a,b)$ the graph of $z=f(x,y)$ near the point $P(a,b,f(a,b))$ looks more and more like a non-vertical plane as zoom in closer and closer to $P$. Here non-vertical means not parallel to the $z$-axis.

But I find it easier to look instead at the level curves of the function $f$ in the $xy$-plane.

\begin{onlineOnly}
    \begin{center}
\desmos{lyquf7jmbb}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/lyquf7jmbb}{163: Linear Approximation Level Sets}









\end{document}