\documentclass{ximera}
\title{Bicycle Tracks}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Parameterizing motion in the plane. 
\end{abstract}
\maketitle

\begin{exploration}

The animation below shows the motion
\[
   {\bf p}_b(t) = \langle a\cos (\omega t), b \sin (\omega t)  \rangle \, , \, 0\leq t \leq 2\pi 
\]
of the rear wheel of a bicycle around an ellipse, where the components are measured in meters and time $t$ in seconds.
Here $a$, $b$, and $\omega$ are positive constants.

\begin{onlineOnly}
    \begin{center}
\desmos{cuo1lynesx}{900}{600}
\end{center}
\end{onlineOnly}

Access Desmos interactives through the online version of this desmos activity at
 
\href{https://www.desmos.com/calculator/cuo1lynesx}{163: Bicycle Tracks 1}.

\begin{enumerate}
\item What are the units of $a$, $b$, and $\omega$? How do you know?

\item  What is the meaning of $\omega$?

\item Suppose the bicyle's frame has length $L$ meters ($L$ is the distance between the front and back axles.) Parameterize the motion of the front wheel. Work in general, not with the specific motion given above. That is, express the position ${\bf p}_f(t)$
of the front wheel (really the point where the wheel touches the ground) at time $t$ in terms of the vector ${\bf p}_b(t)$ and $L$.
Enter your parameterization in Line 25 of the worksheet below as a check.

\item

\item Find an expression for the speed $v_b(t)$ of the back wheel (in m/s) at time $t$ seconds. Graph this function.

\item Find an expression for the speed $v_f(t)$ of the front wheel (in m/s) at time $t$ seconds. Graph this function.

\item  Express the speed of the front wheel in terms of the speed of the back wheels and the angle $\theta$ the front wheel makes with the frame.

\item Express the rate of change $dv_f(t)/dt$ in the speed of the front wheel (with respect to time) in terms of the rate of change $dv_b(t)/dt$ in the speed of the rear wheel and the angle $\theta$. Try to interpret this equation geometrically.

\item Graph the derivatives  $dv_b(t)/dt$, $dv_f(t)/dt$, and their ratio $dv_f(t) / dv_b(t)$.



\end{enumerate}
\end{exploration}

\end{document}
