\documentclass{ximera}
\title{The Acceleration Vector, Part 1}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Acceleration and speed. 
\end{abstract}
\maketitle

In everday English we might say something like ``a car is accelerating'' to mean that the car is speeding up. But this is not really correct. Keep in mind that acceleration measures the rate of change of velocity with respect to time. And since velocity is a vector, so is acceleration. Velocity encodes both speed and direction, so acceleration encodes the rates of change of both speed and direction.

This chapter focuses on determining the rate of change (with respect to time) in speed. For this, we'll need both the acceleration and velocity vectors, or really just the angle between these vectors and the magnitude of the acceleration vector.

Before addressing this question, we'll jump up one level and consider an analogous question about position and velocity. But first, try the following question.

\begin{question}  \label{Q34r05r34gtt}
\begin{enumerate}
\item Interpret the meanings of the expressions 
\[ 
      \Big| \frac{d}{dt} \left( {\bf p}(t) \right) \Big|
\]
and
\[
   \frac{d}{dt} \left( |{\bf p}(t)| \right) .
\]
for a motion with position function ${\bf p}(t)$ measured in meters (and $t$ in seconds). Inlude units in your interpretations. Be sure to state clearly whether each is a vector or a scalar.

\item One of the expressions 
\[ 
      \Big| \frac{d}{dt} \left( {\bf v}(t) \right) \Big|
\]
and
\[
   \frac{d}{dt} \left( |{\bf v}(t)| \right) .
\]
is the magnitude (in $m/s^2)$ of the acceleration vector for a motion with  velocity function ${\bf v}(t)$ measured in meters/sec (and $t$ in seconds). Which one? Interpret the meaning of the other expression.
 
\end{enumerate}
\begin{freeResponse}
\end{freeResponse}
\end{question}



\section{Escaping from the Origin} 
This section under construction. Ignore it for now.

\begin{question}  \label{QErerttg}
\begin{align*}
\frac{d}{dt}\Big| {\bf p}(t)  \Big| &=  \frac{d}{dt} \left(\sqrt{x^2+y^2}\right)    &&   \frac{d}{dt}\Big| {\bf p}(t)  \Big| = \frac{d}{dt} \left(  \sqrt{{\bf p} \cdot {\bf p}} \right) \\
 &= \frac{2x \answer{\frac{dx}{dt}} + 2y \answer{\frac{dy}{dt}}}{2\sqrt{x^2 + y^2}}  && \hskip 0.5 in = \frac{{\bf v}\cdot {\bf p} + {\bf p}\cdot {\bf v}}{2\sqrt{{\bf p}\cdot {\bf p}}} \\
&= \frac{x \frac{dx}{dt} + y \frac{dy}{dt}}{\sqrt{x^2+y^2}}  &&  \hskip 0.5 in = \frac{{\bf v}\cdot {\bf p}}{|{\bf p}|} 
\end{align*}
\end{question}



\section{Speed and Acceleration}
We saw in the last chapter that for a circular motion, we can compute the rate of change in speed with respect to time as the scalar projection of the acceleration vector onto the velocity vector. The same is true for any motion and the proof in this more general case turns out to be very quick.

The idea is to forget about the position function. Instead we'll imagine the hodograph and write the velocity function of a motion
in polar form as
\[
    {\bf v} = v \langle \cos \phi, \sin \phi \rangle .
\]
Here $v = | {\bf v}|$ is the speed and $\phi$ the angle from the positive $x$-axis to the velocity vector, measured counterclockwise. 

Now differentiate with respect to time, keeping in mind that both $v$ and $\phi$ are functions of time. Using the product rule, we find that the acceleration vector is
\[
    {\bf a} = \frac{d{\bf v}}{dt} = \frac{dv}{dt} \langle \cos\phi, \sin\phi \rangle + v \frac{d\phi}{dt} \langle -\sin\phi, \cos \phi \rangle .
\] 

The reason for doing this is that the derivative $dv/dt$ is exactly what we want to compute. It is the rate of change of speed with respect to time. To pick off this rate of change we just take the scalar product of the acceleration with the unit vector
\[
   \frac{{\bf v}}{|{\bf v}|} = \langle \cos\phi, \sin \phi \rangle
\]
in the direction of motion. Then
\begin{align*}
{\bf a} \cdot \frac{{\bf v}}{|{\bf v}|} &= {\bf a} \cdot \langle \cos\phi, \sin \phi \rangle \\
 & = \left(     \frac{dv}{dt} \langle \cos\phi, \sin\phi \rangle + v \frac{d\phi}{dt} \langle -\sin\phi, \cos \phi \rangle      \right) \cdot \langle \cos\phi, \sin \phi \rangle \\
                    & = \frac{dv}{dt} .
\end{align*} 
And that's what we wanted to show.

\emph{The rate of change in the speed of a motion with respect to time is the scalar projection of the acceleration vector onto the velocity vector.}

The advantage of this proof is that it is simple. A disadvantage is that it works only for motions in the plane. The same fact holds for motions in space, but we'll skip the proof for now. Let's look at some examples first.

\section{Projectile Motion}
\begin{question} \label{Q:34re45r45r}
The animation below shows a projectile motion launched at time $t=0$ seconds from position ${\bf p}(0) = \overrightarrow{OA}$ relative to the origin with initial velocity ${\bf v}_0 = {\bf v}(0)$. The projectile moves with constant acceleration ${\bf g}$.

The animation also shows two graphs. One shows speed as a function of time. The other shows the derivative $dv(t)/dt$ of speed with respect to time. 

\begin{enumerate}
\item Identify the two graphs in the animation. Which shows speed as a function of time? Which shows the derivative $dv(t)/dt$ of speed with respect to time?

\item Find an expression for the velocity function ${\bf v}(t)$ using a definite integral. Then evaluate the integral.

\item Find an expression for the position function ${\bf p}(t)$ using a definite integral. Then evaluate the integral.

\item Use the specific vectors ${\bf v}_0$ and ${\bf g}$ below for the initial position, the initial velocity, and the acceleration for the projectile motion in the questions that follow. Assume SI units. 

%\item Use part (b) to find expressions for the position, velocity, and acceleration functions.

\item Find the speed of the motion at time $t=2$ seconds.

\item Is the projectile speeding up or slowing down at time $t=2$ seconds? At what rate. Check your results with the graphs below. 

\item Find an expression for the speed function $v(t)$ and evaluate the limit
\[
    \lim_{t\to \infty} v^\prime (t) .
\]

\item How can you visualize the above limit in the two graphs below?

\item How is the above limit related to the acceleration vector? Explain.

\item Identify the vector ${\bf w}$ in the demonstration above as a vector projection. What role does it play in some of the questions above?

\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\geogebra{vbnga9rj}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.geogebra.org/classic/vbnga9rj}{163: Projectile Motion Plane}

\end{question}


\begin{question}  \label{Q454rft4t4t}
The animation below shows a projectile motion with position function
\[
    {\bf p}(t) = {\bf p}(0) + {\bf v}(0) + \frac{1}{2}{\bf g}t^2 \, , \, t \in \mathbb{R} .
\]
Since the acceleration vector does not point downward, you can think of a small mass sliding on a plane without friction in a uniform gravititational field where the accerelaration does point downward (in the direction of the negative $z$-axis.)

Work in general for this problem. Do \emph{not} make specific choices for the vectors ${\bf p}(0)$, ${\bf v}(0)$, and ${\bf g}$. 

\begin{onlineOnly}
    \begin{center}
\desmosThreeD{clr4mdmltt}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/3d/clr4mdmltt}{163: Projectile 2}

\begin{enumerate}
\item Over what time interval is the projectile slowing down? Speeding up?

\item Assume the trajectory is a parabola and express the position of the vertex relative to the origin in terms of the vectors ${\bf p}(0)$, ${\bf v}(0)$, and ${\bf g}$. Check  your work in the animation above by inputting this position vector on the first blank line.

\item Parameterize the axis of symmetry of the trajectory. Input the parameterization in the worksheet above.

\item Prove algebraically that the trajectory lies in a plane. Find a vector normal to that plane.

\begin{hint}
First find a vector ${\bf n}$ normal to the plane of the trajectory by using two of the three given vectors. Then show that the velocity of the motion is always perpendicular to ${\bf n}$.
\end{hint}


\item Write an equation of the plane of the trajectory. Input your equation in the worksheet above.

\end{enumerate}


\end{question}

\end{document}